---
title: "class.imbalance"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE) 
library(tidyverse)
library(rpart)
library(rpart.plot) 
```

```{r}
library(readxl)
library(dplyr)
fetal.dat <- read_excel("../Data/raw.fetal.data.xlsx")
fetal.dat <- fetal.dat[-1,] #empty first row
#each row is its own id
#variables we'll actually use (including NSP)
#removing date, id, filename, FEB, b, e, AC, DR, SUSP, class, and other classification variables not in our list
fetal <- fetal.dat |> 
  dplyr::select(-c("FileName","Date","SegFile","b", "e", "LBE", "AC", "DR", "A", 
                   "B", "C", "D", "E", "AD","DE","LD","FS","SUSP","CLASS"))
#make NSP a factor (only categorical variable in our selected features)
fetal$NSP <- factor(fetal$NSP)

#2 rows missing data on every variable: can remove
#last row missing some variables, doesn't have a classification label
fetal <- fetal[-c(2127,2128),]
dim(fetal) #2127 people, 20 features, 1 outcome variable

# drop the 1 observation with NA for NSP -> 2126 observations 
fetal <- fetal |> drop_na(NSP) 
```

```{r}
# row is true class, column is predicted class

# all misclassifications are punished equally 
loss_default <- matrix(c(0, 1, 1,
                         1, 0, 1,
                         1, 1, 0), nrow = 3, byrow = TRUE)

# we can change these values, like idk is a false negative 4x as bad as a false positive 
loss_mild <- matrix(c(0, 1, 1,
                         3, 0, 1,
                         4, 2, 0), nrow = 3, byrow = TRUE)

loss_aggressive <- matrix(c(0, 1, 1,
                         12, 0, 1,
                         12, 3, 0), nrow = 3, byrow = TRUE)

calculate_loss <- function(true_labels, predicted_labels, loss_matrix) {
  true_labels <- as.numeric(as.character(true_labels))
  predicted_labels <- as.numeric(as.character(predicted_labels))
  mean(mapply(FUN = function(t, p) loss_matrix[t, p], true_labels, predicted_labels))
}
```

We use synthetic minority oversampling technique (SMOTE) since it requires continuous features and we only use continuous features for our analysis. We only oversample on the training set because we don't want the model to predict the majority class most of the time. We don't use SMOTE on the test set because this represents the proportion of fetal heart statuses we would actually see in clinical settings. 

SMOTE works by creating synthetic points within each minority class. Within each minority class, for each observation A, choose K nearest neighbors and randomly pick one of these neighbors B. Interpolate feature values between observations A and B to create the synthetic observation C. We choose 5 neighbors so synthetic points can be more diverse in terms of feature values. We set dup_size = 0 which lets SMOTE automatically decide how synthetic samples to create in order to create a balanced sample.
```{r}
library(caret)
#install.packages("smotefamily")
library(smotefamily)
set.seed(263)

train_indices <- createDataPartition(fetal$NSP, p = 0.8, list = FALSE)

fetal_train <- fetal[train_indices, ]
fetal_test  <- fetal[-train_indices, ]

# oversample the minority class only on the training set
fetal_balanced_train <- SMOTE(
  X = fetal_train[, -which(names(fetal_train) == "NSP")],
  target = fetal_train$NSP,
  K = 5,   
  dup_size = 0
)$data 

# rename "class" back to "NSP"
names(fetal_balanced_train)[names(fetal_balanced_train) == "class"] <- "NSP"

fetal_balanced_train$NSP <- as.factor(fetal_balanced_train$NSP)
```

Modelling a single tree on an unbalanced training set results in a test accuracy of %92.47. The model misclassifies 2s as 1s (40% of the time).

Modelling a single tree on a balanced training set results in a test accuracy of %91.76 and it still misclassifies 2 as 1s (40% of the time).
```{r}
set.seed(263)

basic_tree_balanced <- rpart(NSP ~ . , data = fetal_balanced_train, method = "class")

# training confusion matrix
basic_tree_balanced_train_prediction <- predict(basic_tree_balanced, fetal_balanced_train, type = "class")

confusionMatrix(basic_tree_balanced_train_prediction, fetal_balanced_train$NSP) # 93.28 train acc

# test confusion matrix 
basic_tree_balanced_test_prediction<-predict(basic_tree_balanced, fetal_test, type = "class")
confusionMatrix(basic_tree_balanced_test_prediction, fetal_test$NSP) # 91.76 test acc

# model predicts a lot of 2's as 1's, so we should punish this more 

# training losses
calculate_loss(basic_tree_balanced_train_prediction, fetal_balanced_train$NSP, loss_default) 
calculate_loss(basic_tree_balanced_train_prediction, fetal_balanced_train$NSP, loss_mild) 
calculate_loss(basic_tree_balanced_train_prediction, fetal_balanced_train$NSP, loss_aggressive)

# test losses
calculate_loss(basic_tree_balanced_test_prediction, fetal_test$NSP, loss_default) 
calculate_loss(basic_tree_balanced_test_prediction, fetal_test$NSP, loss_mild) 
calculate_loss(basic_tree_balanced_test_prediction, fetal_test$NSP, loss_aggressive) 
```
Comparing the test sensitivity computed from the single tree modelled on an unbalanced training set compared to a balanced set, we notice the following: 
The sensitivity stays the same for class 2 and improves slightly from 91.4% to 94.3% for class 3.

Now let's use the balanced data for random forest and gradient boosting modelling, using cross-validation to choose the best hyperparameters.

Remark: For cross validation, we split k-folds on the unbalanced data 'fetal_train', so the validation set doesn't contain synthetic observations and result in an optimistic loss. We want the validation loss to be reflective of what the test loss would likely be. We only apply SMOTE to the training fold and fit our models on the balanced data. We assess the loss (default, mild, aggressive) on the validation fold.

Random forest hyperparameters: mtry (# of predictors to use at given split), ntree
Gradient boosting machine (GBM): n.trees, interaction.depth, shrinkage
```{r}
library(randomForest)
set.seed(263)
# 5 folds
K <- 5 
fold <- sample(1:K, nrow(fetal_train), replace = TRUE)

mtry_values <- c(3, 5, 7, 9) # we have 20 variables, the default is p/3
ntree_values <- c(250, 500, 750, 1000, 1500) # default is 500

rf_cv_results <- data.frame(mtry = c(), ntree = c(), loss_default = c(), loss_mild = c(), loss_aggressive = c())

for(m in mtry_values) {
  for(n in ntree_values) {
    fold_loss_default <- c()
    fold_loss_mild <- c()
    fold_loss_aggressive <- c()
    
    for(k in 1:K) {
      #1. split unbalanced data into train and validation sets
      train_fold <- fetal_train[fold != k, ]
      val_fold <- fetal_train[fold == k, ]
      #2. apply SMOTE only to the training fold
      balanced_train <- SMOTE(
        X = train_fold[, -which(names(train_fold) == "NSP")],
        target = train_fold$NSP,
        K = 5,  # can increase if we want more neighbors 
        dup_size = 0)$data 
      # rename "class" back to "NSP"
      names(balanced_train)[names(balanced_train) == "class"] <- "NSP"
      # ensure categorical outcome is a factor
      balanced_train$NSP <- as.factor(balanced_train$NSP)
      #3. train model on balanced training fold
      rf_model <- randomForest(NSP ~ .,
                               data = balanced_train, 
                               mtry = m,
                               ntree = n)
      #4. assess the loss on the unbalanced validation fold
      val_preds <- predict(rf_model, val_fold)
      fold_loss_default[k] <-
        calculate_loss(val_fold$NSP,
                       val_preds,
                       loss_default)
      fold_loss_mild[k] <- 
        calculate_loss(val_fold$NSP,
                       val_preds,
                       loss_mild)
      fold_loss_aggressive[k] <-
        calculate_loss(val_fold$NSP,
                       val_preds,
                       loss_aggressive)
    }
    
    rf_cv_results <- rf_cv_results |> rbind(data.frame(
      mtry = m,
      ntree = n,
      loss_default = mean(fold_loss_default),
      loss_mild = mean(fold_loss_mild),
      loss_aggressive = mean(fold_loss_aggressive))
    )
    
  }
}

rf_cv_results
```

Repeat a similar process for GBM
```{r warning = FALSE}
library(gbm)
set.seed(263)
get_gbm_labels <- function(gbm_model, dataset, n.trees = 100) { # default n.trees for gbm is 100
  as.factor(max.col((predict.gbm(gbm_model, dataset, type = "response", n.trees = n.trees)[,,1])))
}
gbm_cv_results <- data.frame(n.trees = c(), interaction.depth = c(), shrinkage = c(), loss_default = c(), loss_mild = c(), loss_aggressive = c()) 

n.trees_values <- c(50, 100, 250, 500) # default is 100
interaction.depth_values <- c(1, 2, 3, 4, 5) # default is a depth of 1
shrinkage_values <- c(0.01, 0.1, 0.2) # default is 0.1

for(n in n.trees_values) {
  for(d in interaction.depth_values) {
    for(s in shrinkage_values) {
      fold_loss_default <- c()
      fold_loss_mild <- c()
      fold_loss_aggressive <- c()
      
      for(k in 1:K) {
      #1. split unbalanced data into train and validation sets
      train_fold <- fetal_train[fold != k, ]
      val_fold <- fetal_train[fold == k, ]
      #2. apply SMOTE only to the training fold
      balanced_train <- SMOTE(
        X = train_fold[, -which(names(train_fold) == "NSP")],
        target = train_fold$NSP,
        K = 5,  # can increase if we want more neighbors 
        dup_size = 0)$data 
      # rename "class" back to "NSP"
      names(balanced_train)[names(balanced_train) == "class"] <- "NSP"
      # ensure categorical outcome is a factor
      balanced_train$NSP <- as.factor(balanced_train$NSP)
      #3. train model on balanced training fold
      gbm_model <- gbm(NSP ~ ., data = balanced_train, distribution = "multinomial",
                         n.trees = n, interaction.depth = d, shrinkage = s)
      #4. assess the loss on the unbalanced validation fold
     val_predicted_label <-
       get_gbm_labels(gbm_model, val_fold, n.trees = n)
 
        fold_loss_default[k] <-
          calculate_loss(val_fold$NSP,
                         val_predicted_label,
                         loss_default)
        fold_loss_mild[k] <-
          calculate_loss(val_fold$NSP,
                         val_predicted_label,
                         loss_mild)
        fold_loss_aggressive[k] <-
          calculate_loss(val_fold$NSP,
                         val_predicted_label,
                         loss_aggressive)
      }
      gbm_cv_results <- gbm_cv_results |>
        rbind(data.frame(
          n.trees = n,
          interaction.depth = d,
          shrinkage = s,
          loss_default = mean(fold_loss_default),
          loss_mild = mean(fold_loss_mild),
          loss_aggressive =
            mean(fold_loss_aggressive)))
   }
  }
}
gbm_cv_results
```

Best RF model with default loss:
mtry = 9
ntree = 1500
test accuracy = 2.6%
```{r}
set.seed(263)
# default loss: best rf
best_rf_default_loss <- rf_cv_results[which.min(rf_cv_results$loss_default), ]

best_rf_default_loss
best_rf_default_loss_ntree <- best_rf_default_loss$ntree
best_rf_default_loss_mtry <- best_rf_default_loss$mtry

# fit best model on the balanced data and assess loss on the test set
best_rf_default_loss_model <- randomForest(NSP ~ ., data = fetal_balanced_train,
                                           ntree = best_rf_default_loss_ntree,
                                           mtry  = best_rf_default_loss_mtry)

best_rf_default_loss_model # OOB error = 2.6%

# balanced train
calculate_loss(fetal_balanced_train$NSP, predict(best_rf_default_loss_model, fetal_balanced_train), loss_default) #0.0003214401
# test
calculate_loss(fetal_test$NSP, predict(best_rf_default_loss_model, fetal_test), loss_default) # 0.06117647 (lowest loss out of the three loss models)
```

Best gbm model with default loss:
n.trees = 50
interaction.depth = 5
shrinkage = 0.2
test accuracy = 93.65%
```{r}
# default loss: best gbm
set.seed(263)
best_gbm_default_loss <- gbm_cv_results[which.min(gbm_cv_results$loss_default), ]
best_gbm_default_loss
best_gbm_default_loss_n.trees <- best_gbm_default_loss$n.trees
best_gbm_default_loss_interaction.depth <- best_gbm_default_loss$interaction.depth
best_gbm_default_loss_shrinkage <- best_gbm_default_loss$shrinkage


best_gbm_default_loss_model <- gbm(NSP ~., data = fetal_balanced_train, 
                                   n.trees = best_gbm_default_loss_n.trees, 
                                   interaction.depth = best_gbm_default_loss_interaction.depth,
                                   shrinkage = best_gbm_default_loss_shrinkage)

confusionMatrix(get_gbm_labels(best_gbm_default_loss_model, fetal_balanced_train, n.trees = best_gbm_default_loss_n.trees), fetal_balanced_train$NSP) # balanced train acc = %99.23

confusionMatrix(get_gbm_labels(best_gbm_default_loss_model, fetal_test, n.trees = best_gbm_default_loss_n.trees), fetal_test$NSP) # test acc = %93.65


# balanced train
calculate_loss(fetal_balanced_train$NSP, get_gbm_labels(best_gbm_default_loss_model, fetal_balanced_train, n.trees = best_gbm_default_loss_n.trees), loss_default) 
# test
calculate_loss(fetal_test$NSP, get_gbm_labels(best_gbm_default_loss_model, fetal_test, n.trees = best_gbm_default_loss_n.trees), loss_default) 
```
Best RF model with mild loss:
mtry = 9
ntree = 1500
test accuracy = 2.6%
```{r}
set.seed(263)
# mild loss: best rf
best_rf_mild <- rf_cv_results[which.min(rf_cv_results$loss_mild), ]
best_rf_mild
best_rf_mild_ntree <- best_rf_mild$ntree
best_rf_mild_mtry <- best_rf_mild$mtry

# fit best model on the balanced data and assess loss on the test set
best_rf_mild_loss_model <- randomForest(NSP ~ ., data = fetal_balanced_train,
                                           ntree = best_rf_mild_ntree,
                                           mtry  = best_rf_mild_mtry)

best_rf_mild_loss_model # OOB error = 2.6%

# balanced train
calculate_loss(fetal_balanced_train$NSP, predict(best_rf_mild_loss_model, fetal_balanced_train), loss_mild) # 0.0009643202
# test
calculate_loss(fetal_test$NSP, predict(best_rf_mild_loss_model, fetal_test), loss_mild) # 0.1482353
```

Best gbm model with mild loss:
n.trees = 50
interaction.depth = 4
shrinkage = 0.2
test accuracy = 92.47%
```{r}
set.seed(263)
best_gbm_mild_loss <- gbm_cv_results[which.min(gbm_cv_results$loss_mild), ]
best_gbm_mild_loss
best_gbm_mild_loss_n.trees <- best_gbm_mild_loss$n.trees
best_gbm_mild_loss_interaction.depth <- best_gbm_mild_loss$interaction.depth
best_gbm_mild_loss_shrinkage <- best_gbm_mild_loss$shrinkage


best_gbm_mild_loss_model <- gbm(NSP ~., data = fetal_balanced_train, 
                                   n.trees = best_gbm_mild_loss_n.trees, 
                                   interaction.depth = best_gbm_mild_loss_interaction.depth,
                                   shrinkage = best_gbm_mild_loss_shrinkage)


confusionMatrix(get_gbm_labels(best_gbm_mild_loss_model, fetal_balanced_train, n.trees = best_gbm_mild_loss_n.trees), fetal_balanced_train$NSP) # balanced train acc = 98.87%

confusionMatrix(get_gbm_labels(best_gbm_mild_loss_model, fetal_test, n.trees = best_gbm_mild_loss_n.trees), fetal_test$NSP) # test acc = 92.47%

# balanced train
calculate_loss(fetal_balanced_train$NSP, get_gbm_labels(best_gbm_mild_loss_model, fetal_balanced_train, n.trees = best_gbm_mild_loss_n.trees), loss_mild) 
# test
calculate_loss(fetal_test$NSP, get_gbm_labels(best_gbm_mild_loss_model, fetal_test, n.trees = best_gbm_mild_loss_n.trees), loss_mild) 
```
Best RF model with aggressive loss:
mtry = 9
ntree = 1500
test accuracy = 2.6%
```{r}
set.seed(263)
# aggressive loss: best rf
best_rf_aggressive_loss <- rf_cv_results[which.min(rf_cv_results$loss_aggressive), ]
best_rf_aggressive_loss
best_rf_aggressive_loss_ntree <- best_rf_aggressive_loss$ntree
best_rf_aggressive_loss_mtry <- best_rf_aggressive_loss$mtry

# fit best model on the balanced data and assess loss on the test set
best_rf_aggressive_loss_model <- randomForest(NSP ~ ., data = fetal_balanced_train,
                                           ntree = best_rf_aggressive_loss_ntree,
                                           mtry  = best_rf_aggressive_loss_mtry)

best_rf_aggressive_loss_model # OOB error = 2.6% (same as the other two loss model)

# balanced train
calculate_loss(fetal_balanced_train$NSP, predict(best_rf_aggressive_loss_model, fetal_balanced_train), loss_aggressive) # 0.003857281
# test
calculate_loss(fetal_test$NSP, predict(best_rf_aggressive_loss_model, fetal_test), loss_aggressive) # 0.5058824
```
Best gbm model with aggressive loss:
n.trees = 50
interaction.depth = 3
shrinkage = 0.2
test accuracy = 94.59%
```{r}
set.seed(263)
best_gbm_aggressive_loss <- gbm_cv_results[which.min(gbm_cv_results$loss_aggressive), ]
best_gbm_aggressive_loss
best_gbm_aggressive_loss_n.trees <- best_gbm_aggressive_loss$n.trees
best_gbm_aggressive_loss_interaction.depth <- best_gbm_aggressive_loss$interaction.depth
best_gbm_aggressive_loss_shrinkage <- best_gbm_aggressive_loss$shrinkage


best_gbm_aggressive_loss_model <- gbm(NSP ~., data = fetal_balanced_train, 
                                   n.trees = best_gbm_aggressive_loss_n.trees, 
                                   interaction.depth = best_gbm_aggressive_loss_interaction.depth,
                                   shrinkage = best_gbm_aggressive_loss_shrinkage)

confusionMatrix(get_gbm_labels(best_gbm_aggressive_loss_model, fetal_balanced_train, n.trees = best_gbm_aggressive_loss_n.trees), fetal_balanced_train$NSP) # balanced train acc = 99.97%

confusionMatrix(get_gbm_labels(best_gbm_aggressive_loss_model, fetal_test, n.trees = best_gbm_aggressive_loss_n.trees), fetal_test$NSP) # test acc = 94.59%

# balanced train
calculate_loss(fetal_balanced_train$NSP, get_gbm_labels(best_gbm_aggressive_loss_model, fetal_balanced_train, n.trees = best_gbm_aggressive_loss_n.trees), loss_aggressive) 
# test
calculate_loss(fetal_test$NSP, get_gbm_labels(best_gbm_aggressive_loss_model, fetal_test, n.trees = best_gbm_aggressive_loss_n.trees), loss_aggressive) 
```

