---
title: "263 Project"
author: "Hannah Hamling"
date: "2026-04-30"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE) 
library(tidyverse)
```

## Read in data/variable selection
```{r, message = FALSE}
library(readxl)
library(dplyr)
fetal.dat <- read_excel("../Data/raw.fetal.data.xlsx")
fetal.dat <- fetal.dat[-1,] #empty first row
#each row is its own id
#variables we'll actually use (including NSP)
#removing date, id, filename, FEB, b, e, AC, DR, SUSP, class, and other classification variables not in our list
fetal <- fetal.dat |> 
  dplyr::select(-c("FileName","Date","SegFile","b", "e", "LBE", "AC", "DR", "A", 
                   "B", "C", "D", "E", "AD","DE","LD","FS","SUSP","CLASS"))
#make NSP a factor (only categorical variable in our selected features)
fetal$NSP <- factor(fetal$NSP)

dim(fetal) #2129 rows (people), 20 features, 1 outcome variable (NSP)
#2 rows missing data on every variable: can remove
#last row missing some variables, doesn't have a classification label
fetal <- fetal[-c(2127,2128),]
dim(fetal) #2127 people, 20 features, 1 outcome variable
fetal[!complete.cases(fetal), ] #only this one row missing data
```

## Look at class frequencies (for later class minority adjustment)
```{r}
summary(fetal$NSP)
#1655 are normal, 295 are suspect, 176 are pathologic, 1 missing
#plot(fetal$NSP)

# drop the 1 observation with NA for NSP -> 2126 observations 
fetal <- fetal |> drop_na(NSP) 

#1655 are normal, 295 are suspect, 176 are pathologic
plot(fetal$NSP, main = "Fetal Heart Rate Levels", sub = "Figure 1")

#install.packages("flextable")
library(flextable)
library(dplyr)
as.data.frame(prop.table(table(fetal$NSP))) |>
  rename("NSP Level" = Var1, "Proportion" = Freq) |>
  flextable() |>
  set_caption("Figure 2: Proportions of Fetal Heart Rate Levels")
```

```{r}
library(caret)
set.seed(263)

#80/20 split
train_indices <- createDataPartition(fetal$NSP, p = 0.8, list = FALSE)

fetal_train <- fetal[train_indices, ]
fetal_test  <- fetal[-train_indices, ]

# make sure there are some of each NSP in both train and test data 
table(fetal_train$NSP)
table(fetal_test$NSP)
```

# Custom Asymmetric Loss Analysis

We have a classification problem with 3 classes of fetal state (1 = Normal, 2 = Suspect, 3 = Pathological). Classes 2 and 3 are underrepresented compared to class 1. Clinically, certain misclassifications are more harmful than others. 

* Mistaking 3 as 1: Worst possible mistake 
* Mistaking 2 as 1: Also a pretty bad mistake 
* Mistaking 3 as 2: Bad, but not as bad as the previous two cases
* Mistaking 2 as 3: not that bad
* Mistaking 1 as 2: not that bad 
* Mistaking 1 as 3: not that bad 

Thus, we can define a custom loss that punishes certain mixups accordingly.

```{r}
# row is true class, column is predicted class

# all misclassifications are punished equally 
loss_default <- matrix(c(0, 1, 1,
                         1, 0, 1,
                         1, 1, 0), nrow = 3, byrow = TRUE)

# we can change these values, like idk is a false negative 4x as bad as a false positive 
#4 times as bad
loss_mild <- matrix(c(0, 1, 1,
                         3, 0, 1,
                         4, 2, 0), nrow = 3, byrow = TRUE)

#12 times as bad
loss_aggressive <- matrix(c(0, 1, 1,
                         12, 0, 1,
                         12, 3, 0), nrow = 3, byrow = TRUE)

calculate_loss <- function(true_labels, predicted_labels, loss_matrix) {
  true_labels <- as.numeric(true_labels)
  predicted_labels <- as.numeric(predicted_labels)
  mean(mapply(FUN = function(t, p) loss_matrix[t, p], true_labels, predicted_labels))
}
```

## Default Single Tree Using All Variables

Let's see how a single basic tree with all variables and default arguments does with our custom loss. We see that 2 gets misclassified as 1 a lot (bad mistake), which increases the loss.

```{r}
library(rpart)
library(rpart.plot) 

set.seed(263)

basic_tree <- rpart(NSP ~ . , data = fetal_train, method = "class")

#rpart.plot(basic_tree)

# training confusion matrix
basic_tree_train_prediction <- predict(basic_tree, fetal_train, type = "class")
confusionMatrix(basic_tree_train_prediction, fetal_train$NSP) # 93.36 train acc

# test confusion matrix 
basic_tree_test_prediction <- predict(basic_tree, fetal_test, type = "class")
confusionMatrix(basic_tree_test_prediction, fetal_test$NSP) # 92.47 test acc

# model predicts a lot of 2's as 1's, so we should punish this more 

# training losses: highest with aggressive
calculate_loss(basic_tree_train_prediction, fetal_train$NSP, loss_default) #0.06643151
calculate_loss(basic_tree_train_prediction, fetal_train$NSP, loss_mild) #0.09817754
calculate_loss(basic_tree_train_prediction, fetal_train$NSP, loss_aggressive) #0.2016461

# test losses: highest with aggressive
calculate_loss(basic_tree_test_prediction, fetal_test$NSP, loss_default) #0.07529412
calculate_loss(basic_tree_test_prediction, fetal_test$NSP, loss_mild) #0.09882353
calculate_loss(basic_tree_test_prediction, fetal_test$NSP, loss_aggressive) #0.2047059
```

## Default RF

The default RF model shows a large class error rate for class 2 (and 3 somewhat), so the loss increases when we use our custom loss.

```{r}
library(randomForest)
set.seed(263)
rf_default <- randomForest(NSP ~ ., data = fetal_train)
rf_default # once again, 2 is being misclassified as 1, OOB error = 5.7%

# train losses: highest with aggressive loss
calculate_loss(predict(rf_default, fetal_train), fetal_train$NSP, loss_default) #0.001175779
calculate_loss(predict(rf_default, fetal_train), fetal_train$NSP, loss_mild) #0.002351558
calculate_loss(predict(rf_default, fetal_train), fetal_train$NSP, loss_aggressive) #0.007642563

# test losses: highest with aggressive loss
calculate_loss(predict(rf_default, fetal_test), fetal_test$NSP, loss_default) #0.05882353
calculate_loss(predict(rf_default, fetal_test), fetal_test$NSP, loss_mild) #0.08470588
calculate_loss(predict(rf_default, fetal_test), fetal_test$NSP, loss_aggressive) #0.1929412
```

## Default Boosting 

Boosting model also misclassifies 2 as 1 a lot. 3 as 1 too sometimes.

```{r warning = FALSE}
library(gbm)

get_gbm_labels <- function(gbm_model, dataset, n.trees = 100) { # default n.trees for gbm is 100
  as.factor(max.col((predict.gbm(gbm_model, dataset, type = "response", n.trees = n.trees)[,,1])))
}

set.seed(263)
gbm_default <- gbm(NSP ~ ., data = fetal_train, distribution = "multinomial")

# training confusion matrix
gbm_default_train_prediction <- get_gbm_labels(gbm_default, fetal_train)
confusionMatrix(gbm_default_train_prediction, fetal_train$NSP) # 94.42 train acc

# test confusion matrix 
gbm_default_test_prediction <- get_gbm_labels(gbm_default, fetal_test)
confusionMatrix(gbm_default_test_prediction, fetal_test$NSP) # 92.47 test acc 

# train losses: highest with aggressive
calculate_loss(gbm_default_train_prediction, fetal_train$NSP, loss_default) #0.0558495
calculate_loss(gbm_default_train_prediction, fetal_train$NSP, loss_mild) #0.09112287
calculate_loss(gbm_default_train_prediction, fetal_train$NSP, loss_aggressive) #0.2186949

# test losses: highest with aggressive
calculate_loss(gbm_default_test_prediction, fetal_test$NSP, loss_default) #0.07529412
calculate_loss(gbm_default_test_prediction, fetal_test$NSP, loss_mild) #0.1176471
calculate_loss(gbm_default_test_prediction, fetal_test$NSP, loss_aggressive) #0.3082353
```


## RF CV 

now doing cv to select the best RF hyperparameters but according to our custom loss (hopefully helps to fix the bad sorts of mixups)

Hyperparameters include: mtry and ntree

```{r}
set.seed(263)
# 5 folds
K <- 5 
fold <- sample(1:K, nrow(fetal_train), replace = TRUE)

mtry_values <- c(3, 5, 7, 9) # we have 20 variables, the default is p/3
ntree_values <- c(250, 500, 750, 1000, 1500) # default is 500

rf_cv_results <- data.frame(mtry = c(), ntree = c(), loss_default = c(), loss_mild = c(), loss_aggressive = c())

for(m in mtry_values) {
  for(n in ntree_values) {
    fold_loss_default <- c()
    fold_loss_mild <- c()
    fold_loss_aggressive <- c()
    
    for(k in 1:K) {
      train <- fetal_train[fold != k, ]
      val <- fetal_train[fold == k, ]
      
      rf_model <- randomForest(NSP ~ .,
                               data = train, 
                               mtry = m,
                               ntree = n)
      
      fold_loss_default[k] <- calculate_loss(val$NSP, predict(rf_model, val), loss_default)
      fold_loss_mild[k] <- calculate_loss(val$NSP, predict(rf_model, val), loss_mild)
      fold_loss_aggressive[k] <- calculate_loss(val$NSP, predict(rf_model, val), loss_aggressive)
    }
    
    rf_cv_results <- rf_cv_results |> rbind(data.frame(
      mtry = m,
      ntree = n,
      loss_default = mean(fold_loss_default),
      loss_mild = mean(fold_loss_mild),
      loss_aggressive = mean(fold_loss_aggressive))
    )
    
  }
}

rf_cv_results
which.min(rf_cv_results$loss_default) #10
rf_cv_results[10,] #mtry of 5, 1500 trees
which.min(rf_cv_results$loss_mild) #13
rf_cv_results[13,] #mtry of 7, 750 trees
which.min(rf_cv_results$loss_aggressive) #13
rf_cv_results[13,] #mtry of 7, 750 trees
```


## GBM CV 

now doing cv to select the best GBM hyperparameters but according to our custom loss (hopefully helps to fix the bad sorts of mixups)

Hyperparameters include: n.trees, interaction.depth, shrinkage

```{r warning = FALSE, message = FALSE}
set.seed(263)
gbm_cv_results <- data.frame(n.trees = c(), interaction.depth = c(), shrinkage = c(), loss_default = c(), loss_mild = c(), loss_aggressive = c()) 

n.trees_values <- c(50, 100, 250, 500) # default is 100
interaction.depth_values <- c(1, 2, 3, 4, 5) # default is a depth of 1
shrinkage_values <- c(0.01, 0.1, 0.2) # default is 0.1

for(n in n.trees_values) {
  for(d in interaction.depth_values) {
    for(s in shrinkage_values) {
      fold_loss_default <- c()
      fold_loss_mild <- c()
      fold_loss_aggressive <- c()
      
      for(k in 1:K) {
        train <- fetal_train[fold != k, ]
        val <- fetal_train[fold == k, ]
        
        gbm_model <- gbm(NSP ~ ., data = train, distribution = "multinomial",
                         n.trees = n, interaction.depth = d, shrinkage = s)
        
        
        val_predicted_label <- get_gbm_labels(gbm_model, val, n.trees = n)
 
        fold_loss_default[k] <- calculate_loss(val$NSP, val_predicted_label, loss_default)
        fold_loss_mild[k] <- calculate_loss(val$NSP, val_predicted_label, loss_mild)
        fold_loss_aggressive[k] <- calculate_loss(val$NSP, val_predicted_label, loss_aggressive)
      }
    gbm_cv_results <- gbm_cv_results |> rbind(data.frame(
      n.trees = n,
      interaction.depth = d,
      shrinkage = s,
      loss_default = mean(fold_loss_default),
      loss_mild = mean(fold_loss_mild),
      loss_aggressive = mean(fold_loss_aggressive))
    )
    }
  }
}

gbm_cv_results 

which.min(gbm_cv_results$loss_default) #44
gbm_cv_results[44,] #default best at 250 trees, depth of 5, shrinkage of 0.1

which.min(gbm_cv_results$loss_mild) #41
gbm_cv_results[41,] #default best at 250 trees, depth of 4, shrinkage of 0.1

which.min(gbm_cv_results$loss_aggressive) #42
gbm_cv_results[42,] #default best at 250 trees, depth of 4, shrinkage of 0.2
```

## Choosing Best Model based on Custom Loss

We choose to sacrifice overall accuracy in order to improve in positive identification of classes 2 and 3.

### Default Loss (equal punishment for all misclassifications)

Best RF model with default loss is: ntree = 1500, mtry = 5

```{r}
set.seed(263)
best_rf_default_loss <- rf_cv_results[which.min(rf_cv_results$loss_default), ]
best_rf_default_loss
best_rf_default_loss_ntree <- best_rf_default_loss$ntree
best_rf_default_loss_mtry <- best_rf_default_loss$mtry
best_rf_default_loss_model <- randomForest(NSP ~., data = fetal_train, 
                                           ntree = best_rf_default_loss_ntree, 
                                           mtry = best_rf_default_loss_mtry)
best_rf_default_loss_model # OOB error = 5.29%, lots of 2's classified as 1's
#class 1 error = 0.01057402
#class 2 error = 0.26694915, labels as 1's quite a bit
#class 3 error = 0.09219858

calculate_loss(fetal_train$NSP, predict(best_rf_default_loss_model, fetal_train), loss_default) #0.001175779
calculate_loss(fetal_test$NSP, predict(best_rf_default_loss_model, fetal_test), loss_default) #0.06117647
```

Best GBM with default loss is n.trees = 250, interaction.depth = 5, shrinkage = 0.1

```{r}
set.seed(263)
best_gbm_default_loss <- gbm_cv_results[which.min(gbm_cv_results$loss_default), ]
best_gbm_default_loss
best_gbm_default_loss_n.trees <- best_gbm_default_loss$n.trees
best_gbm_default_loss_interaction.depth <- best_gbm_default_loss$interaction.depth
best_gbm_default_loss_shrinkage <- best_gbm_default_loss$shrinkage
best_gbm_default_loss_model <- gbm(NSP ~., data = fetal_train, 
                                   n.trees = best_gbm_default_loss_n.trees, 
                                   interaction.depth = best_gbm_default_loss_interaction.depth,
                                   shrinkage = best_gbm_default_loss_shrinkage)

confusionMatrix(get_gbm_labels(best_gbm_default_loss_model, fetal_train, n.trees = best_gbm_default_loss_n.trees), fetal_train$NSP) # train acc = 99.84%

confusionMatrix(get_gbm_labels(best_gbm_default_loss_model, fetal_test, n.trees = best_gbm_default_loss_n.trees), fetal_test$NSP) # test acc = 93.65%
#class 1 balanced accuracy = 0.9051
#class 2 balanced accuaracy = 0.8748
#class 3 balanced accuracy = 0.94158

calculate_loss(fetal_train$NSP, get_gbm_labels(best_gbm_default_loss_model, fetal_train, n.trees = best_gbm_default_loss_n.trees), loss_default) #0.0005878895
calculate_loss(fetal_test$NSP, get_gbm_labels(best_gbm_default_loss_model, fetal_test, n.trees = best_gbm_default_loss_n.trees), loss_default) #0.06352941
```

### Mild Loss

Best RF model with mild loss is: ntree = 750, mtry = 7

```{r}
set.seed(263)
best_rf_mild_loss <- rf_cv_results[which.min(rf_cv_results$loss_mild), ]
best_rf_mild_loss
best_rf_mild_loss_ntree <- best_rf_mild_loss$ntree
best_rf_mild_loss_mtry <- best_rf_mild_loss$mtry
best_rf_mild_loss_model <- randomForest(NSP ~., data = fetal_train, 
                                        ntree = best_rf_mild_loss_ntree, 
                                        mtry = best_rf_mild_loss_mtry)
best_rf_mild_loss_model # OOB error = 5.47%, still pretty bad at 2's but better at 3's than default loss RF
#class 1 error = 0.01283988
#class 2 error = 0.27542373
#class 3 error = 0.07801418

calculate_loss(fetal_train$NSP, predict(best_rf_mild_loss_model, fetal_train), loss_mild) #0.0005878895
calculate_loss(fetal_test$NSP, predict(best_rf_mild_loss_model, fetal_test), loss_mild) #0.1576471
```

Best GBM with mild loss is n.trees = 250, interaction.depth = 4, shrinkage = 0.1

```{r}
set.seed(263)
best_gbm_mild_loss <- gbm_cv_results[which.min(gbm_cv_results$loss_mild), ]
best_gbm_mild_loss
best_gbm_mild_loss_n.trees <- best_gbm_mild_loss$n.trees
best_gbm_mild_loss_interaction.depth <- best_gbm_mild_loss$interaction.depth
best_gbm_mild_loss_shrinkage <- best_gbm_mild_loss$shrinkage
best_gbm_mild_loss_model <- gbm(NSP ~., data = fetal_train, 
                                   n.trees = best_gbm_mild_loss_n.trees, 
                                   interaction.depth = best_gbm_mild_loss_interaction.depth,
                                   shrinkage = best_gbm_mild_loss_shrinkage)

confusionMatrix(get_gbm_labels(best_gbm_mild_loss_model, fetal_train, n.trees = best_gbm_mild_loss_n.trees), fetal_train$NSP) # train acc = 99.88%

confusionMatrix(get_gbm_labels(best_gbm_mild_loss_model, fetal_test, n.trees = best_gbm_mild_loss_n.trees), fetal_test$NSP) # test acc = 94.82%
#balanced class accuracies: 0.9173   0.8762  0.93901

calculate_loss(fetal_train$NSP, get_gbm_labels(best_gbm_mild_loss_model, fetal_train, n.trees = best_gbm_mild_loss_n.trees), loss_mild) #0.001175779
calculate_loss(fetal_test$NSP, get_gbm_labels(best_gbm_mild_loss_model, fetal_test, n.trees = best_gbm_mild_loss_n.trees), loss_mild) #0.1317647
```

### Aggressive Loss 

Best RF model with aggressive loss is: ntree = 750, mtry = 7, same as mild loss best RF hyperparameters 

```{r}
set.seed(263)
best_rf_aggressive_loss <- rf_cv_results[which.min(rf_cv_results$loss_aggressive), ]
best_rf_aggressive_loss
best_rf_aggressive_loss_ntree <- best_rf_aggressive_loss$ntree
best_rf_aggressive_loss_mtry <- best_rf_aggressive_loss$mtry
best_rf_aggressive_loss_model <- randomForest(NSP ~., data = fetal_train, 
                                        ntree = best_rf_aggressive_loss_ntree, 
                                        mtry = best_rf_aggressive_loss_mtry)
best_rf_aggressive_loss_model # OOB error = 5.47%, still pretty bad at 2's but better at 3's than default loss RF
  #class 1 error = 0.01283988
  #class 2 error = 0.27542373 (flags a lot as 1's)
  #class 3 error = 0.07801418 (some FN)

calculate_loss(fetal_train$NSP, predict(best_rf_aggressive_loss_model, fetal_train), loss_aggressive) #0.0005878895
calculate_loss(fetal_test$NSP, predict(best_rf_aggressive_loss_model, fetal_test), loss_aggressive) #0.5576471
```

Best GBM with aggressive loss is n.trees = 250, interaction.depth = 4, shrinkage = 0.2. This model has the most balanced accuracy 

```{r}
set.seed(263)
best_gbm_aggressive_loss <- gbm_cv_results[which.min(gbm_cv_results$loss_aggressive), ]
best_gbm_aggressive_loss #250 trees, depth of 4, shrankge of 0.2
#loss aggressive = 0.4320668
best_gbm_aggressive_loss_n.trees <- best_gbm_aggressive_loss$n.trees
best_gbm_aggressive_loss_interaction.depth <- best_gbm_aggressive_loss$interaction.depth
best_gbm_aggressive_loss_shrinkage <- best_gbm_aggressive_loss$shrinkage
best_gbm_aggressive_loss_model <- gbm(NSP ~., data = fetal_train, 
                                   n.trees = best_gbm_aggressive_loss_n.trees, 
                                   interaction.depth = best_gbm_aggressive_loss_interaction.depth,
                                   shrinkage = best_gbm_aggressive_loss_shrinkage)

confusionMatrix(get_gbm_labels(best_gbm_aggressive_loss_model, fetal_train, n.trees = best_gbm_aggressive_loss_n.trees), fetal_train$NSP) # train acc = 99.84%
#only incorrectly classified one 'normal' as suspect

confusionMatrix(get_gbm_labels(best_gbm_aggressive_loss_model, fetal_test, n.trees = best_gbm_aggressive_loss_n.trees), fetal_test$NSP) # test acc = 94.59%
#class balanced accuracies 0.9279   0.9016  0.94029

calculate_loss(fetal_train$NSP, get_gbm_labels(best_gbm_aggressive_loss_model, fetal_train, n.trees = best_gbm_aggressive_loss_n.trees), loss_aggressive) #0.0005878895
calculate_loss(fetal_test$NSP, get_gbm_labels(best_gbm_aggressive_loss_model, fetal_test, n.trees = best_gbm_aggressive_loss_n.trees), loss_aggressive) #0.3482353
```