---
title: "263 Project"
author: "Hannah Hamling"
date: "2026-04-30"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Read in data/variable selection
```{r, message = FALSE}
library(readxl)
library(dplyr)
fetal.dat <- read_excel("../Data/raw.fetal.data.xlsx")
fetal.dat <- fetal.dat[-1,] #empty first row
#each row is its own id
#variables we'll actually use (including NSP)
#removing date, id, filename, FEB, b, e, AC, DR, SUSP, class, and other classification variables not in our list
fetal <- fetal.dat |> 
  dplyr::select(-c("FileName","Date","SegFile","b", "e", "LBE", "AC", "DR", "A", 
                   "B", "C", "D", "E", "AD","DE","LD","FS","SUSP","CLASS"))
#make NSP a factor (only categorical variable in our selected features)
fetal$NSP <- factor(fetal$NSP)

dim(fetal) #2129 rows (people), 20 features, 1 outcome variable (NSP)
#2 rows missing data on every variable: can remove
#last row missing some variables, doesn't have a classification label
fetal <- fetal[-c(2127,2128),]
dim(fetal) #2127 people, 20 features, 1 outcome variable
fetal[!complete.cases(fetal), ] #only this one row missing data


```

## Look at class frequencies (for later class minority adjustment)
```{r}
summary(fetal$NSP)
#1655 are normal, 295 are suspect, 176 are pathologic, 1 missing
#plot(fetal$NSP)

# drop the 1 observation with NA for NSP -> 2126 observations 
fetal <- fetal |> drop_na(NSP) 

#1655 are normal, 295 are suspect, 176 are pathologic
plot(fetal$NSP)
```

```{r}
library(caret)
set.seed(263)

train_indices <- createDataPartition(fetal$NSP, p = 0.8, list = FALSE)

fetal_train <- fetal[train_indices, ]
fetal_test  <- fetal[-train_indices, ]

# make sure there are some of each NSP in both train and test data 
table(fetal_train$NSP)
table(fetal_test$NSP)
```

# Basic Tree Using All Variables

```{r}
library(rpart)
library(rpart.plot) 

basic_tree <- rpart(NSP ~ . , data = fetal_train, method = "class")

rpart.plot(basic_tree)

# training confusion matrix
basic_tree_train_prediction <- predict(basic_tree, fetal_train, type = "class")
confusionMatrix(basic_tree_train_prediction, fetal_train$NSP) # 93.36 train acc

# test confusion matrix 
basic_tree_test_prediction <- predict(basic_tree, fetal_test, type = "class")
confusionMatrix(basic_tree_test_prediction, fetal_test$NSP) # 92.47 test acc

# model predicts a lot of 2's as 1's, so we should punish this more 
```

## Basic Tree with Harsh Punishment for Common Mixup 
```{r}
# punish the common mistake of classifying a 2 as a 1

# row is true class, column is predicted class
# harsh punishment for misclassifying 1 as 2 (# can do like cv or something for this value)
punishment <- 2
loss_matrix_2misclassified1 <- matrix(c(
  0,  1,  1,  
  punishment,  0,  1,  # harsh punishment for misclassifying 2 as 1
  1,  1,  0 ), nrow = 3, byrow = TRUE)
rownames(loss_matrix_2misclassified1) <- c("class1", "class2", "class3")
colnames(loss_matrix_2misclassified1) <- c("class1", "class2", "class3")

# control the tree by requiring at least 5% of the training data in each node
basic_tree_punish_2misclass1 <- rpart(NSP ~ . , data = fetal_train, method = "class",
                               parms = list(loss = loss_matrix_2misclassified1))
rpart.plot(basic_tree_punish_2misclass1)

# training confusion matrix
basic_tree_punish_2misclass1_train_prediction <- predict(basic_tree_punish_2misclass1, fetal_train, type = "class")
confusionMatrix(basic_tree_punish_2misclass1_train_prediction, fetal_train$NSP) # train acc = .9406

# test confusion matrix 
basic_tree_punish_2misclass1_test_prediction <- predict(basic_tree_punish_2misclass1, fetal_test, type = "class")
confusionMatrix(basic_tree_punish_2misclass1_test_prediction, fetal_test$NSP) # test acc = 0.9176 

# better at getting 2's 


```


## Common Tree with Lower Prob Threshold to Classifying Class 2

```{r}
basic_tree_train_predicted_probs <- predict(basic_tree, fetal_train, type = "prob")
basic_tree_test_predicted_probs <- predict(basic_tree, fetal_test, type = "prob")
prob_threshold_class2 <- 0.2 # can tune this with cv 

basic_tree_train_prediction_threshold <- as.factor(ifelse(basic_tree_train_predicted_probs[, 2] > prob_threshold_class2, "2", colnames(basic_tree_train_predicted_probs)[max.col(basic_tree_train_predicted_probs)]))
confusionMatrix(basic_tree_train_prediction_threshold, fetal_train$NSP)

basic_tree_test_prediction_threshold <- as.factor(ifelse(basic_tree_test_predicted_probs[, 2] > prob_threshold_class2, "2", colnames(basic_tree_test_predicted_probs)[max.col(basic_tree_test_predicted_probs)]))
confusionMatrix(basic_tree_test_prediction_threshold, fetal_test$NSP)
```

# Bagging, Random Forest
```{r}
library(randomForest)

set.seed(263)
rf_default <- randomForest(NSP ~ ., data = fetal_train)
rf_default # once again, 2 is being misclassified as 1 
```

## RF with higher weight for class 2, all default params
```{r}
set.seed(263)
rf_weight2 <- 1000000 # can tune w cv 
rf_weight2 <- randomForest(NSP ~ ., data = fetal_train, classwt = c(`1` = 1, `2` = rf_weight2, `3` = 1))
rf_weight2
```

## RF with lower probability threshold to predict 2

```{r}
set.seed(263)
threshold2 <- 0.2 # can tune w cv 
rf_threshold2 <- randomForest(NSP ~ ., data = fetal_train, cutoff = c((1 - threshold2) / 2, threshold2, (1 - threshold2) / 2))
rf_threshold2
```

# Boosting, defaults 

```{r}
library(gbm)
set.seed(263)
gbm_default <- gbm(NSP ~ ., data = fetal_train)

# training confusion matrix
gbm_default_train_prediction <- as.factor(max.col((predict.gbm(gbm_default, fetal_train, type = "response")[,,1])))
confusionMatrix(gbm_default_train_prediction, fetal_train$NSP) # 94.42 train acc

# test confusion matrix 
gbm_default_test_prediction <- as.factor(max.col((predict.gbm(gbm_default, fetal_test, type = "response")[,,1])))
confusionMatrix(gbm_default_test_prediction, fetal_test$NSP) # 92.47 test acc 

```

## Boosting, higher weight for class 2 errors

```{r}
set.seed(263)
gbm_weight2_magnitude <- 3
gbm_weight <- ifelse(fetal_train$NSP == 2, gbm_weight2_magnitude, 1)
gbm_weight2 <- gbm(NSP ~ ., data = fetal_train, weights = gbm_weight)

# training confusion matrix
gbm_weight2_train_prediction <- as.factor(max.col((predict.gbm(gbm_weight2, fetal_train, type = "response")[,,1])))
confusionMatrix(gbm_weight2_train_prediction, fetal_train$NSP) # 93.3 train acc

# test confusion matrix 
gbm_weight2_test_prediction <- as.factor(max.col((predict.gbm(gbm_weight2, fetal_test, type = "response")[,,1])))
confusionMatrix(gbm_weight2_test_prediction, fetal_test$NSP) # 92.94 test acc 

# reduces the misclassification of 2 as 1
```

## Boosting, lowered probability threshold for class 2

```{r}
# using the default boosting with even weights for all classes 

gmb_2threshold <- 0.27

gbm_default_train_probs <- as.data.frame(predict.gbm(gbm_default, fetal_train, type = "response"))
colnames(gbm_default_train_probs) <- c("1", "2", "3")
gbm_threshold_train_predictions <- as.factor(ifelse(gbm_default_train_probs$`2` > gmb_2threshold, "2",
               colnames(gbm_default_train_probs)[max.col(gbm_default_train_probs)]))

gbm_default_test_probs <- as.data.frame(predict.gbm(gbm_default, fetal_test, type = "response"))
colnames(gbm_default_test_probs) <- c("1", "2", "3")
gbm_threshold_test_predictions <- as.factor(ifelse(gbm_default_test_probs$`2` > gmb_2threshold, "2",
               colnames(gbm_default_test_probs)[max.col(gbm_default_test_probs)]))

# training confusion matrix
confusionMatrix(gbm_threshold_train_predictions, fetal_train$NSP) # 92.59 train acc

# test confusion matrix 
confusionMatrix(gbm_threshold_test_predictions, fetal_test$NSP) # 92 test acc 
```


