---
title: "263 Project"
author: "Hannah Hamling"
date: "2026-04-30"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


## Read in data/variable selection
```{r, message = FALSE}
library(readxl)
library(dplyr)
fetal.dat <- read_excel("../Data/raw.fetal.data.xlsx")
fetal.dat <- fetal.dat[-1,] #empty first row
#each row is its own id
#variables we'll actually use (including NSP)
#removing date, id, filename, FEB, b, e, AC, DR, SUSP, class, and other classification variables not in our list
fetal <- fetal.dat |> 
  dplyr::select(-c("FileName","Date","SegFile","b", "e", "LBE", "AC", "DR", "A", 
                   "B", "C", "D", "E", "AD","DE","LD","FS","SUSP","CLASS"))
#make NSP a factor (only categorical variable in our selected features)
fetal$NSP <- factor(fetal$NSP)

dim(fetal) #2129 rows (people), 20 features, 1 outcome variable (NSP)
#2 rows missing data on every variable: can remove
#last row missing some variables, doesn't have a classification label
fetal <- fetal[-c(2127,2128),]
dim(fetal) #2127 people, 20 features, 1 outcome variable
fetal[!complete.cases(fetal), ] #only this one row missing data


```

## Look at class frequencies (for later class minority adjustment)
```{r}
summary(fetal$NSP)
#1655 are normal, 295 are suspect, 176 are pathologic, 1 missing
#plot(fetal$NSP)

# drop the 1 observation with NA for NSP -> 2126 observations 
fetal <- fetal |> drop_na(NSP) 

#1655 are normal, 295 are suspect, 176 are pathologic
plot(fetal$NSP)
```

```{r}
library(caret)
set.seed(263)

train_indices <- createDataPartition(fetal$NSP, p = 0.8, list = FALSE)

fetal_train <- fetal[train_indices, ]
fetal_test  <- fetal[-train_indices, ]

# make sure there are some of each NSP in both train and test data 
table(fetal_train$NSP)
table(fetal_test$NSP)
```

# Custom Asymmetric Loss Analysis

We have a classification problem with 3 classes of fetal state (1 = Normal, 2 = Suspect, 3 = Pathological). Classes 2 and 3 are underrepresented compared to class 1. Clinically, certain misclassifications are more harmful than others. 

* Mistaking 3 as 1: Worst possible mistake 
* Mistaking 2 as 1: Also a pretty bad mistake 
* Mistaking 3 as 2: Bad, but not as bad as the previous two cases
* Mistaking 2 as 3: not that bad
* Mistaking 1 as 2: not that bad 
* Mistaking 1 as 3: not that bad 

Thus, we can define a custom loss that punishes certain mixups accordingly.

```{r}
# row is true class, column is predicted class

# all misclassifications are punished equally 
loss_default <- matrix(c(0, 1, 1,
                         1, 0, 1,
                         1, 1, 0), nrow = 3, byrow = TRUE)

# we can change these values, like idk is a false negative 5x as bad as a false positive 
loss_mild <- matrix(c(0, 1, 1,
                         4, 0, 1,
                         5, 2, 0), nrow = 3, byrow = TRUE)


loss_aggressive <- matrix(c(0, 1, 1,
                         7, 0, 1,
                         10, 2, 0), nrow = 3, byrow = TRUE)

calculate_loss <- function(true_labels, predicted_labels, loss_matrix) {
  true_labels <- as.numeric(true_labels)
  predicted_labels <- as.numeric(predicted_labels)
  mean(mapply(FUN = function(t, p) loss_matrix[t, p], true_labels, predicted_labels))
}
```



## Default Single Tree Using All Variables

Let's see how a single basic tree with all variables and default arguments does with our custom loss. We see that 2 gets misclassified as 1 a lot (bad mistake), which increases the loss.

```{r}
library(rpart)
library(rpart.plot) 

set.seed(263)

basic_tree <- rpart(NSP ~ . , data = fetal_train, method = "class")

#rpart.plot(basic_tree)

# training confusion matrix
basic_tree_train_prediction <- predict(basic_tree, fetal_train, type = "class")
confusionMatrix(basic_tree_train_prediction, fetal_train$NSP) # 93.36 train acc

# test confusion matrix 
basic_tree_test_prediction <- predict(basic_tree, fetal_test, type = "class")
confusionMatrix(basic_tree_test_prediction, fetal_test$NSP) # 92.47 test acc

# model predicts a lot of 2's as 1's, so we should punish this more 

# training losses
calculate_loss(basic_tree_train_prediction, fetal_train$NSP, loss_default) # default, 0.0664
calculate_loss(basic_tree_train_prediction, fetal_train$NSP, loss_mild) # mild, 0.11
calculate_loss(basic_tree_train_prediction, fetal_train$NSP, loss_aggressive) # aggressive, 0.1558

# test losses
calculate_loss(basic_tree_test_prediction, fetal_test$NSP, loss_default) # default, 0.075
calculate_loss(basic_tree_test_prediction, fetal_test$NSP, loss_mild) # mild, 0.111
calculate_loss(basic_tree_test_prediction, fetal_test$NSP, loss_aggressive) # aggressive, 0.146
```

## Default RF

The default RF model shows a large class error rate for class 2 (and 3 somewhat), so the loss increases when we use our custom loss.

```{r}
library(randomForest)
set.seed(263)
rf_default <- randomForest(NSP ~ ., data = fetal_train)
rf_default # once again, 2 is being misclassified as 1, OOB error = 5.7%

# train losses
calculate_loss(predict(rf_default, fetal_train), fetal_train$NSP, loss_default) # default, 0.00118
calculate_loss(predict(rf_default, fetal_train), fetal_train$NSP, loss_mild) # mild, 0.00294
calculate_loss(predict(rf_default, fetal_train), fetal_train$NSP, loss_aggressive) # aggressive, 0.0047

# test losses
calculate_loss(predict(rf_default, fetal_test), fetal_test$NSP, loss_default) # default, 0.0588
calculate_loss(predict(rf_default, fetal_test), fetal_test$NSP, loss_mild) # mild, 0.0964
calculate_loss(predict(rf_default, fetal_test), fetal_test$NSP, loss_aggressive) # aggressive, 0.132
```

## Default Boosting 

Boosting model also misclassifies 2 as 1 a lot. 3 as 1 too sometimes.

```{r}
library(gbm)
set.seed(263)
gbm_default <- gbm(NSP ~ ., data = fetal_train)

# training confusion matrix
gbm_default_train_prediction <- as.factor(max.col((predict.gbm(gbm_default, fetal_train, type = "response")[,,1])))
confusionMatrix(gbm_default_train_prediction, fetal_train$NSP) # 94.42 train acc

# test confusion matrix 
gbm_default_test_prediction <- as.factor(max.col((predict.gbm(gbm_default, fetal_test, type = "response")[,,1])))
confusionMatrix(gbm_default_test_prediction, fetal_test$NSP) # 92.47 test acc 

# train losses 
calculate_loss(gbm_default_train_prediction, fetal_train$NSP, loss_default) # default, 0.0558
calculate_loss(gbm_default_train_prediction, fetal_train$NSP, loss_mild) # mild, 0.1058
calculate_loss(gbm_default_train_prediction, fetal_train$NSP, loss_aggressive) # aggressive, 0.1605

# test losses
calculate_loss(gbm_default_test_prediction, fetal_test$NSP, loss_default) # default, 0.0752
calculate_loss(gbm_default_test_prediction, fetal_test$NSP, loss_mild) # mild, 0.1388
calculate_loss(gbm_default_test_prediction, fetal_test$NSP, loss_aggressive) # aggressive, 0.202
```


## RF CV 

now doing cv to select the best RF hyperparameters but according to our custom loss (hopefully helps to fix the bad sorts of mixups)

Hyperparameters include: mtry and ntree

```{r}
# warning this chunk takes TIME 
set.seed(263)
# 5 folds
K <- 5 
fold <- sample(1:K, nrow(fetal_train), replace = TRUE)

mtry_values <- c(3, 5, 7, 9) # we have 20 variables, the default is p/3
ntree_values <- c(250, 500, 750, 1000, 1500) # default is 500

rf_cv_results <- data.frame(mtry = c(), ntree = c(), loss_default = c(), loss_mild = c(), loss_aggressive = c())

for(m in mtry_values) {
  for(n in ntree_values) {
    fold_loss_default <- c()
    fold_loss_mild <- c()
    fold_loss_aggressive <- c()
    
    for(k in 1:K) {
      train <- fetal_train[fold != k, ]
      val <- fetal_train[fold == k, ]
      
      rf_model <- randomForest(NSP ~ .,
                               data = train, 
                               mtry = m,
                               ntree = n)
      
      fold_loss_default[k] <- calculate_loss(val$NSP, predict(rf_model, val), loss_default)
      fold_loss_mild[k] <- calculate_loss(val$NSP, predict(rf_model, val), loss_mild)
      fold_loss_aggressive[k] <- calculate_loss(val$NSP, predict(rf_model, val), loss_aggressive)
    }
    
    rf_cv_results <- rf_cv_results |> rbind(data.frame(
      mtry = m,
      ntree = n,
      loss_default = mean(fold_loss_default),
      loss_mild = mean(fold_loss_mild),
      loss_aggressive = mean(fold_loss_aggressive))
    )
    
  }
}

rf_cv_results
```


```{r}
# this is the best set of hyperparameters for the mild loss
# we see a lower class error in classes 2 and 3 compared to the default og RF model
# still kinda bad for class 2 tho 
randomForest(NSP ~., data = fetal_train, ntree = 750, mtry = 7)
```



## GBM CV 

now doing cv to select the best GBM hyperparameters but according to our custom loss (hopefully helps to fix the bad sorts of mixups)

Hyperparameters include: n.trees, interaction.depth, shrinkage

```{r warning = FALSE, message = FALSE}
gbm_cv_results <- data.frame(n.trees = c(), interaction.depth = c(), shrinkage = c(), loss_default = c(), loss_mild = c(), loss_aggressive = c())

n.trees_values <- c(50, 100, 250, 500, 1000) # default is 100
interaction.depth_values <- c(1, 2, 3) # default is a depth of 1
shrinkage_values <- c(0.001, 0.01, 0.1, 0.2) # default is 0.1

for(n in n.trees_values) {
  for(d in interaction.depth_values) {
    for(s in shrinkage_values) {
      fold_loss_default <- c()
      fold_loss_mild <- c()
      fold_loss_aggressive <- c()
      
      for(k in 1:K) {
        train <- fetal_train[fold != k, ]
        val <- fetal_train[fold == k, ]
        
        gbm_model <- gbm(NSP ~ ., data = train, distribution = "multinomial",
                         n.trees = n, interaction.depth = d, shrinkage = s)
        
        val_probs <- predict(gbm_model, val, n.trees = n, type = "response")[,,1]
        colnames(val_probs) <- c("1", "2", "3")
        val_predicted_label <- colnames(val_probs)[max.col(val_probs)]
 
        fold_loss_default[k] <- calculate_loss(val$NSP, val_predicted_label, loss_default)
        fold_loss_mild[k] <- calculate_loss(val$NSP, val_predicted_label, loss_mild)
        fold_loss_aggressive[k] <- calculate_loss(val$NSP, val_predicted_label, loss_aggressive)
      }
    gbm_cv_results <- gbm_cv_results |> rbind(data.frame(
      n.trees = n,
      interaction.depth = d,
      shrinkage = s,
      loss_default = mean(fold_loss_default),
      loss_mild = mean(fold_loss_mild),
      loss_aggressive = mean(fold_loss_aggressive))
    )
    }
  }
}

gbm_cv_results
```

```{r}
# best hyperparameters according to loss mild is 250 trees, depth 3, 0.2 shrinkage 
set.seed(263)
gbm1 <- gbm(NSP ~., data = fetal_train, n.trees = 250, interaction.depth = 3, shrinkage = 0.2)
gbm1_probs_train <- predict(gbm1, fetal_train, n.trees = n, type = "response")[,,1]
colnames(gbm1_probs_train) <- c("1", "2", "3")
gbm1_class_train <- as.factor(colnames(gbm1_probs_train)[max.col(gbm1_probs_train)])
confusionMatrix(gbm1_class_train, fetal_train$NSP) 

gbm1_probs_test <- predict(gbm1, fetal_test, n.trees = n, type = "response")[,,1]
colnames(gbm1_probs_test) <- c("1", "2", "3")
gbm1_class_test <- as.factor(colnames(gbm1_probs_test)[max.col(gbm1_probs_test)])
confusionMatrix(gbm1_class_test, fetal_test$NSP) # OH WOW it really did good on classes 2 and 3 compared to before 
```